---
title: "STA221"
author: "Neil Montgomery"
date: "Last edited: `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: 
  beamer_presentation:
    keep_tex: TRUE
    incremental: TRUE
#    df_print: tibble
    fig_caption: FALSE
classoption: aspectratio=169
header-includes:
- \newcommand{\ve}{\varepsilon}
- \newcommand{\dbar}[1]{\overline{\overline{#1}}}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE,
                      dev='pdf', fig.width=5, fig.asp=0.618, fig.align='center')
options(tibble.width=70, tibble.print_min=5, show.signif.stars = FALSE)
library(tidyverse)
```

```{r}
burn <- read.delim("chelating.txt", check.names = FALSE, 
                   colClasses = c("factor", "factor", "numeric"))
```



## the analysis

We have new sums of squares decompositions. The details can get brutal - so we'll stay symbolic. 

We'll call the factor variables `A` and `B`, with $I$ and $J$ levels respectively.

In the additive case we get, assuming a common sample size of $n$ for each combination of factor levels (strongly encouraged in practice!):

$$SS_{Total} =  SS_A + SS_B + SS_{Error}$$

\pause Everything has $\chi^2$ distributions. The degrees of freedom add up (N is the grand sample size):

$$N-1 = (I-1) + (J-1) + (N-I-J+1)$$

\pause When the experiment is \textit{balanced} (strongly encouraged!) with $n$ in each combination of levels, the error degrees of freedom simplifies to $IJ(n-1)$.

## fire example - no interaction (?!)

```{r}
burn_fit <- burn %>% 
  aov(`Burn-Times` ~ Bath + Launderings, data=.)
summary(burn_fit)
```

## sums of squares - a few details

The total sum of squares $SS_{Total}$ is (as always) $(N-1)$ times the sample varianace of the response variable:
$$\sum_{i,j,k} \left(y_{ijk} - \dbar{y}\right)^2$$

The treatment sums of squares will be:
$$SS_A = nI\sum_i\left(\overline{y}_{i\cdot\cdot} - \dbar{y}\right)^2$$

$$SS_B = nJ\sum_j\left(\overline{y}_{\cdot j \cdot} - \dbar{y}\right)^2$$

where the dots in the subscript mean "averaged over this index."

\pause The key thing to note is that each treatment sum of squares is computed using the average over all levels of the other treatment...

\pause ...which only makes sense when there is no interaction.

## error sum of squares

For the sake of completeness:

$$SS_{Error} = \sum_{i,j,k} \left(y_{ijk} -\overline{y}_{ij\cdot}\right)^2$$

Note that $y_{ijk} -\overline{y}_{ij\cdot}$ is also called a "residual".

## model assumptions

Mostly the same as with one treatment factor, with the same verification techniques.

1. Observations are independent (this is assumed - only knowledge of the experiment itself is of any help in satisfying this assumption.)

2. Equal variance for all combinations of levels of treatment factors. (Plots, or Levene's test when $n$ is not too small.) (Fatal if violated.)

3. Normal error (normal quantile plot of residuals) (Violation OK if sample size is large enough.)

\pause When $n=1$, the lack of interaction is also an *assumption*. 

## fire retardant model assumptions - equal variance

Plot of residuals versus "fitted values" (in this case, just the group averages):

```{r}
library(broom)
augment(burn_fit) %>% 
  ggplot(aes(x=.fitted, y=.resid)) + geom_jitter(width = 0.1, height=0.1)
```

## fire retardant model assumptions - equal variance

Since $n=12$ Levene's test also works:

 
```{r}
library(car)
burn %>% leveneTest(`Burn-Times` ~ Launderings*Bath, data=.)
```

## fire retardant model assumptions - normality

```{r}
augment(burn_fit) %>% 
  ggplot(aes(sample=.resid)) + geom_qq()
```

## the general model and analysis (with interaction)

This model has the $(\tau\gamma)$ interaction term:

$$y_{ijk} = \mu + \tau_i + \gamma_j + (\tau\gamma)_{ij} +  \ve_{ijk} \qquad \text{ general model}$$

\pause The sum of squares decomposition is now:

$$SS_{Total} = SS_{A} + SS_{B} + SS_{AB} + SS_{Error}$$

\pause The interaction sum of squares is:
$$n\sum_{i,j}\left(y_{ijk} - \overline{y}_{i\cdot\cdot} - \overline{y}_{\cdot j \cdot} + \dbar{y}\right)^2$$

Small when additive; large when not.

## degrees of freedom - balanced case

\begin{align*}
SS_{Total} &= SS_{A} + SS_{B} + SS_{AB} + SS_{Error}\\
(N-1) &= (I-1) + (J-1) + (I-1)(J-1) + IJ(n-1)
\end{align*}

Note: $IJ(n-1) = N - IJ$

We get (in addition):

$$\frac{SS_{AB}/(I-1)(J-1)}{SS_{Error}/IJ(n-1)} \sim F_{(I-1)(J-1), IJ(n-1)}$$

If there is evidence for interaction, do not try to interpret the "main effects". 

## flame retardant with interaction

Flame retardant example, with interaction (and without):

```{r}
burn_fit_int <- burn %>%
  aov(`Burn-Times` ~ Bath*Launderings, data=.)
summary(burn_fit_int)
summary(burn_fit)  
```

## unreplicated designs

If $n=1$, it is possible to proceed with the analysis, \textit{if you can assume there is no interaction.}

\pause It isn't possible to test for interaction when $n=1$. In principle, because there is no variation within each combination to estimate.

\pause Also, the ANOVA table wouldn't work!

\pause The technique is to use $SS_{AB}$ as the error sum of squares.

## another overall example

Chromatography example. 

Two factors: flow rate (fast and slow); Concentration (low, med, high)

## interaction plot

```{r}
ch <- read.csv("Chromatography.csv")
with(ch, interaction.plot(Concentration, Flow.Rate, Counts))
```


## analysis

```{r}
ch_fit <- ch %>% 
  aov(Counts ~ Flow.Rate*Concentration, data=.) 
summary(ch_fit)
  
```

## assumptions

```{r}
ch %>% leveneTest(Counts ~ Concentration*Flow.Rate, data=.)
```

## assumptions

```{r}
library(broom)
augment(ch_fit) %>% 
  ggplot(aes(sample=.resid)) + geom_qq()
```


